**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**

### 2019-01-23 submission/manuscript.Rmd
add 2nd par

Advances in sequencing technology and decreasing costs of generating multitudes of 16S rRNA gene sequences allowed rapid exploration and taxonomic characterization of the human associated microbiome. Currently, the microbiome field is growing at an unprecedented rate and with it, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health. Human associated microbial communities are remarkably uneven and it is unlikely that a single species can explain a disease state universally. It is more likely that subsets of those communities, in relation to one another, reliably account for the unevenness and be used as biomarkers of a specific disease state. Thus, researchers have started to explore the utility of machine learning (ML) techniques to identify the biomarkers associated with healty and diseased individuals. However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are reproducible. Moreover, there is a lack of consideration on why a particular method is utilized.

>>>>

Advances in sequencing technology and decreasing costs of generating 16S rRNA gene sequences allowed rapid exploration and taxonomic characterization of the human associated microbiome. Currently, the microbiome field is growing at an unprecedented rate and as a result, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and human health. Human associated microbial communities are remarkably uneven and it is unlikely that a single species can explain a disease state comprehensively. It is more likely that subsets of those communities, in relation to one another and to their host, can reliably account for the unevenness and be used as biomarkers of a specific disease state. Thus, researchers have started to explore the utility of machine learning (ML) techniques to identify the biomarkers associated with healthy and diseased individuals. However, currently the field's use of ML lacks clarity and consistency on which methods are used, how these methods are implemented and if these methods are reproducible. Moreover, there is a lack of consideration on why a particular method is utilized. Often, there is a trade-off between the interpretibility and accuracy of ML methods. As researchers, we have to find the right balance to accomplish complex knowledge tasks such as identifying a disease state based on microbiome-associated biomarkers while also understanding how these tasks are accomplished.

One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers. Colorectal cancer (CRC) is one of the leading causes of death in the US, therefore developing a microbiome assay as a non-invasive screening tool would be a useful innovation. However, the efforts to combine ML methods with relative abundances of bacterial poplations in stool to predict CRC tumors have been afflicted by the aforementioned flawed ML practices. As a result, the predictive performance of these models varies greatly with areas under the receiver operating characteristic curve (AUROC) of 0.7-0.9.
